---
title: "Template for ATPase rate calculation (using rollapply)"
author: "Todo"
date: 'Todo'
output: html_notebook

---

# Background
`rollapply` is a function that allows to apply a function to a rolling window within a data set.
Since finding a good region of the NADH measurements to determine the slope manually is difficult and time intensive, using roll apply provides a fast solution to semi-automate that task.
The function `rollapply` is used to calculate the slope of a rolling window of the NADH measurements.
Afterwards, the region where the slopes display a plateau of minimal values is selected and an average slope is calculated to determine the steepest part of the graph.
The slope of each well is then used to calculate the ATPase rate of the respective sample (well).

```{r setup}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(magrittr)
library(ggplot2)
library(rbokeh)
library(RColorBrewer)
library(openxlsx)
library(zoo)

theme_set(theme_minimal())
theme_update(panel.grid = element_line(colour = "gray88"))
palette("Dark2")
metaData <- list(experimentId = "todo", 
                 LonConc = 0.05,
                 SubstrateRefConc = 2, 
                 comment = "5 mM PEP, 0.2 mM NADH")



###
# define some functions used later
# returns the labels saved in the vector columns based on the given vector of names (= wells)
getLabels <- function(elements) {
  ret <- list()
  
  for (elem in elements) {
    ret %<>% append(values = columns[elem])
  }
  
  return(ret)
}


# returns all wells of a certain sample (using the columns vector)
getWells <- function(sampleName = "blank") {
  ret <- c()
  
  for (element in names(columns)) {
    if (tolower(columns[element]) == tolower(sampleName)) {
      ret %<>% append(values = element)
    }
  }
  
  return(ret)
}

```


# Setup
## samples
<!-- todo: specify samples measured -->
Check (1× = 2µM):

* List of samples

## Conditions

<!-- todo: update conditions -->
* 1× Lon reaction buffer
* 10% Glycerol
* 1 mM DTT
* 2 mM ATP
* 5 mM PEP
* 0.2 mM NADH
* 10 U PK/LDH
* 0.05 µM Lon-His
* x µM Substrate

# Determining ATPase rate

## Load data

The NADH absorption at 340 nm was measured in a Tecan Spark platereader.
In addition, an initial measurement of the absorptions at 900 nm and  1000 nm was conducted at the beginning to normalize the absorption measurments to a path length of 1 cm. 
The measured values are then exported to an Excel file (.xlsx) and the various wavelength are split manually into separate sheets.
That file of measurments is loaded here, "s" and " °C" is removed from columns time and temp, respectively and converted into numbers.

```{r loadData, echo=FALSE}

columns <- c(
  "time" = "time",
  "temp" = "temp",
  "H1" = "todo",
  "H2" = "todo",
  "H3" = "todo",
  "H4" = "todo",
  "H5" = "todo",
  "H6" = "todo",
  "H7" = "todo",
  "H8" = "todo",
  "H9" = "todo",
  "H10" = "todo",
  "H11" = "todo",
  "H12" = "todo"
)
fileLocation <- "../measurement/file.xlsx"

abs340 <- read.xlsx(xlsxFile = fileLocation, sheet = "NADH (340nm)") 
abs1000 <- read.xlsx(xlsxFile = fileLocation, sheet = "TestWL (1000nm)")
abs900 <- read.xlsx(xlsxFile = fileLocation, sheet = "RefWL (900nm)")


abs340$time %<>% gsub(pattern = "s", replacement = "") %>% 
  as.integer()
abs340$temp %<>% gsub(pattern = "°C", replacement = "") %>%
  as.numeric()
  
head(abs340)
abs1000
abs900
```

## Raw data

Raw data as obtained from the machine.

```{r plotRaw, fig.width=10}
abs340 %>% 
  pivot_longer(cols = !time:temp,
               names_to = "sample",
               values_to = "abs340") %>% 
  ggplot(aes(y=abs340, x=time, col=sample)) +
  geom_line() +
  ## uncomment to show points in addition to line
  # geom_point() +
  scale_y_continuous() + 
  # scale_x_continuous(n.breaks = 8) +
  expand_limits(x=c(0, 0.7), y=0) +
  labs(x = "time [s]", y = expression("Absorbance"["340nm"]), title = "Raw data") +
  facet_wrap(vars(sample), labeller = as_labeller(columns))

```


## Normalized data

```{r normalizeData, fig.width=10}
# Test wave length: 1000nm
# Reference wave length: 900
# Measured in a 1 cm cuvette in a spectrophotometer (Vis V-1200 from VWR) using the buffer stated above.
TestWl1cm <- c(0.226, 0.225, 0.224)
RefWl1cm <- c(0.078, 0.077, 0.078)
#calculating the K-factor
kFactor <- mean(TestWl1cm) - mean(RefWl1cm)


# caluclate the path length corrected values
abs340Corrected <- abs340 
dataCols2 <- abs340Corrected  %>%
  select(!time & !temp) %>%
  colnames()
abs340Corrected[dataCols2] <- mapply("*", 
                                 abs340Corrected[dataCols2],
                                 kFactor / (abs1000[dataCols2] - abs900[dataCols2]))


# subtract the blank and remove the blank from the dataset
abs340Corrected %<>% 
  mutate(blank = rowMeans(select(., all_of(getWells("blank"))))) %>%
  pivot_longer(names_to = "well", values_to = "abs340", cols = matches("^[A-H][1-9][012]?$")) %>%
  mutate(abs340 = abs340 - blank) %>%
  filter(!well %in% getWells("blank")) %>% 
  mutate(label = getLabels(well))

# plot normalized data
abs340Corrected %>%
  ggplot(aes(x = time/60, y = abs340, col = well)) +
  geom_line() +
  # Formatting
  scale_y_continuous() +
  scale_x_continuous(n.breaks = 6) +
  expand_limits(x=c(0, 0.7), y=0) +
  # Labels
  labs(x = "time [min]", 
       y = expression("normalized absorbance"["340nm"]),
       title = "normalized Absorption") +
  facet_wrap(~ well, labeller = as_labeller(columns))
```


## Fit and plot slope
The normalized data is then used to fit the slopes within a window of sequential data points.
In case of a continuous degradation the slopes should plateau for a bit until the reaction stops.
At that point the slopes will quickly increase to 0 and stay there.
Usually, a window of 40 is a good window with to smooth out noise, while still capturing the shape of the curve.
In some cases, the reaction runs so fast, that a window of 40 data points spans past the end of the reaction.
In those cases, the window needs to be shorter (20 or even 10).

In a second step, the range containing the plateau in the beginning is chosen and used to calculate the average slope within that range.

```{r fitSlope, fig.width=10}
fittingWindow <- 40
fittingFunction <- function(absorption, time_s) {
  fittingLimit <- fittingWindow/2
  # absorption has the indexes according to the row number --> use those to find the correct times
  timeRange <- absorption %>%
    names() %>%
    as.integer()
  
  if (length(timeRange) <= fittingLimit) {
    ret <- list(c("(Intercept)" = 0, "time_s" = 0))
  } else {
    ret <-
      tibble(absorption = absorption, time_s = time_s[timeRange]) %>%
      lm(formula = absorption ~ time_s)
  }
  
  ret %>%
    return()
}

abs340Corrected %<>% 
  group_by(well) %>%
  mutate(fit = rollapply(abs340, time, width = fittingWindow, partial = TRUE,
                            FUN = fittingFunction,
                            by.column = FALSE, align = "left"
                            ),
         windowWidth = fittingWindow) %>%
   mutate(slope = sapply(fit[,1], function(x){x[2]}))

## In some cases with a quick reaction the window needs to be reduced to 20 or 
## even 10 data points to be able to capture the slope.
## In those cases uncomment this section and do fitting again with shorter window.
# fittingWindow <- 20
# abs340Corrected[abs340Corrected$well %in% c("todo"),] %<>% 
#   group_by(well) %>%
#   mutate(fit = rollapply(abs340, time, width = fittingWindow, partial = TRUE,
#                             FUN = fittingFunction,
#                             by.column = FALSE, align = "left"
#                             ),
#          windowWidth = fittingWindow) %>%
#    mutate(slope = sapply(fit[,1], function(x){x[2]}))

```

```{r AvgSlopeRange, fig.width=10}
###
# Define the range of relevant slopes that should be used to calculate the average Slope
# todo: define ranges for calculation of the average slope
ranges <-  tribble(
  ~well, ~lower, ~upper,
  "H1", 0, 4300,
  "H2", 0, 4300,
  "H3", 0, 4300,
  "H4", 0, 4300,
  "H5", 0, 4300,
  "H6", 0, 4300,
  "H7", 0, 4300,
  "H8", 0, 4300,
  "H9", 0, 4300,
  "H10", 0, 4300,
  "H11", 0, 4300,
  "H12", 0, 4300
)


###
# plot the slope with the selected ranges
# filtering time allows a detailed look at a certain range.
plotSlope <- abs340Corrected %>% 
#  filter(time < 1500) %>%
  ggplot(aes(x = time, y = slope, col = well)) +
  geom_line() +
  expand_limits(y = c(0)) +
  facet_wrap(~well, scale = "free_x") +
  geom_vline(data = ranges,
             mapping = aes(xintercept = lower),
             color = "red",
             size = .1)+
  geom_vline(data = ranges,
             mapping = aes(xintercept = upper),
             color = "red",
             size = .1) +
  labs(x = "time [s]", y = "slope")
plotSlope

plotAbs <- abs340Corrected %>% 
  ggplot(aes(x = time, y = abs340, col = well)) +
  geom_line() +
  expand_limits(y = c(0)) +
  facet_wrap(~well) +
  geom_vline(data = ranges,
             mapping = aes(xintercept = lower),
             color = "red",
             size = .1) +
  geom_vline(data = ranges,
             mapping = aes(xintercept = upper + (fittingWindow * 16)),
             color = "red",
             size = .1) +
  labs(x = "time [s]", y = "normalized absorption [340nm]")
plotAbs

```


## Average Slope

The average slope in the defined region is calculated.
Afterwards, the curve including a horizontal line indicating the average slope is plotted.

```{r calcAvgSlope, fig.width=10}
rates <- abs340Corrected %>%
  left_join(ranges, by = "well") %>%
  filter(lower <= time, time <= upper) %>%
  group_by(well) %>%
  summarise(across(slope, list(mean = mean, median = median), .names = "{.fn}_{.col}"),
            across(windowWidth, list(min = min, max = max), .names = "{.fn}_{.col}"))

plotSlope +
  geom_hline(data = rates,
             mapping = aes(yintercept = mean_slope),
             color = "black",
             size = .2)
```



## Calculating the ATPase rates

Now, use the formula to calculate the ATPase rates:

$$r(ATPase) = \frac{1}{\epsilon(NADH) * c(ATPase) * d} * \frac{d(A{340nm})}{dt}$$

Where:

* $\epsilon(NADH) = 6.220*10^3 M^{-1} cm^{-1}$ This is the extinction coefficient of NADH
* $c(ATPase)$ is the concentration of hexameric Lon used (in M)
* $d = 1cm$ (the path length that we normalized to 1cm)
* $\frac{d(A{340nm})}{dt}$ being the slope of the curve per **min**

In our case the slopes are in s^-1^ and need to be converted to min^-1^

```{r calcATPaseRates}
# conversion factor from s^-1 to min^-1 for the slope
convSL <- 60
# epsilon(NADH) (in M^-1 cm^-1)
eNADH <- 6220
# Lon concentration conversion from µM to M
convLon <- (1 / 1000) / 1000
# path length in cm
pathlength <- 1


rates %<>%
  mutate(ATPaseRate = (-mean_slope) * convSL * 1 / (eNADH * metaData[["LonConc"]] * convLon * pathlength)) %>% 
  mutate(labs = columns[well])


rates %>%
  ggplot(aes(x = well, y = ATPaseRate, fill = labs)) +
  geom_col(color = "black", size = .2)  +
  scale_fill_brewer(palette = "Paired")
```

Finally, we need to first subtract the value of "NADH only" from all others and then the value of "PK/LDH".
Doing that, we get the ATPase rate of the respective sample.

```{r ATPaseRates}

rates %<>% 
  #set experiment number
  mutate(experiment = experimentId,
         comment = metaData[["comment"]],
         Lon.conc = metaData[["LonConc"]]) %>%
  # subtract basal PK/LDH activity from ATPase rates 
  # (as this also includes the basal NADH oxidation rate, subtracting the NADH sample from the others can be skipped)
  mutate(ATPaseRate_pkldh = ATPaseRate - mean(ATPaseRate[.$well %in% getWells(sampleName = "PK/LDH")]))


# plot actual ATPase rates
rates %>%
  ggplot(aes(x = well, y = ATPaseRate_pkldh, fill = labs)) +
  geom_col(color = "black", size = .2) +
  scale_fill_brewer(palette = "Paired")

# select all except NAHD and PK/LDH
rates %>%
  select(labs, well, ATPaseRate_pkldh, Lon.conc, experiment, comment, min_windowWidth) %>%
  filter(!tolower(labs) %in% c("nadh", "pk/ldh")) %>%
  write.xlsx(file = paste0("./out-ATPaseRates-", metaData[["experimentId"]], ".xlsx"), overwrite = TRUE)
```